---
output:
  govdown::govdown_document:
    font: "sans-serif" # (default) or "new-transport"
    favicon: "none" # or "none" (default) or "govuk" see ?govdown_document
    logo: FALSE # or false (default)
    logo_url: "https://ukgovdatascience.github.io/govdown/"
    logo_text: "DHSC C&C"
    page_title: "page_title"
    title: "Using GoogleMaps API to get drive-time distances"
    author: "Lisa Laeber"
    date: "July 2021"
    phase: alpha
    feedback_url: "https://github.com/ukgovdatascience/govdown/issues"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Getting started with Google APIs

Before using any of Google's APIs you'll need to create an API key. 
You will need to provide credit card details to obtain an API key. 
The 'free' plan for the distance matrix api allows up to 2,500 requests within a 
24h window and up to $200 worth of requests per month. 
To find out more about this, check out <https://developers.google.com/maps/documentation/distance-matrix/get-api-key>. 

## Sample use case

For this C&C sessions, we'll try to find the *n* closest care homes and the drive time to get there.

### 1. Set-up: Packages, API key, data, function definitions
```{r}
# 1. packages ----------------------------------------------------------------
pacman::p_load(dplyr, sf, mapsapi, httr, leaflet)

# 2. Read API key from my googleCloud platform project 'DHSC-geo-project-Apr21'
source('R/define_google_api_key.R')

# 3. Read in data about ~15k UK care homes of interest
care_homes <- readxl::read_excel("data/carehome Location Data & project overview.xlsx",
                                 sheet = "input_carehomes") %>% 
  # remove 3 care homes without coordinates
  filter(!(is.na(locationlong)|is.na(locationlat)))
```

Geo data is usually provided in form of coordinates - typically longitude and latitude, but if your data does come in a different format (e.g. a post code) it's best to transform geo data into coordinates before you get started.
Let's define a function to transform a postcode into geo coordinates.

Using the free post *postcodes.io* API, we can transform a postcode into easting/northings or long/lat and obtain a few extra bits of administrative information (see https://postcodes.io/).

NOTE: this could be done through gmaps api's mp_geocode function, too, but let's save those requests for what we really need to have them for.
(If of interest see, the gmaps api way: https://developers.google.com/maps/documentation/geocoding/overview)

```{r}
# Define function
postcode_lookup <- function(postcode) {
  infos_about_postcode <- httr::GET(paste0("https://api.postcodes.io/postcodes/", postcode))
  httr::warn_for_status(infos_about_postcode)
  httr::content(infos_about_postcode)
}

```

### 2. Specifiying our 'user' parameters

Now, let's define a starting point, e.g. a random address, and say how many care homes to display
```{r}
# DHSC address: Quarry House, Quarry Hill, Leeds, West Yorkshire Ls2 7Ue, Leeds LS2 7UE
my_location_name <- "DHSC office in Leeds: Quarry House, Quarry Hill, Leeds LS2 7UE"
my_postcode <- "LS2 7UE"

top_n <- 20

```

### 3. Transforming data into the right format

Extract coordinate information and transform into geo object:
```{r}
# returns a list of information associated with post code
infos_about_my_postcode <- postcode_lookup(my_postcode)

my_address_df <- data.frame(my_long = infos_about_my_postcode$result$longitude, 
                         my_lat = infos_about_my_postcode$result$latitude)

my_address_sf <- my_address_df %>% 
  sf::st_as_sf(coords = c("my_long", "my_lat"), crs = 4326) 

```


### 4. Calculations
We do the distance calculations in two steps to reduce the number of requests to the API:

#### 4.1. Pre-calculate straight-line distances
Pre-calculate straight-line distance from our address to all ~15k care homes and keep top n care homes.
```{r}
# use sf package to first pre-calculate approximate distance. (was geosphere pkg)
carehomes_distances <- care_homes %>% 
  sf::st_as_sf(coords = c("locationlong", "locationlat"), crs = 4326) %>% 
  sf::st_distance(my_address_sf)
  
# returns a distance matrix
head(carehomes_distances)
```

```{r}
# Ensure that query sticks to google API limit of max 25 destination per start point
max_elements <- 25 
max_n <- min(top_n, max_elements)

# add distances to df and keep only top n closest (by straight-line dist) care homes
top_n_carehomes_df <- care_homes %>% 
  select(CQClocationid, locationname, locationlat, locationlong) %>% 
  cbind(Distance = carehomes_distances) %>% 
  arrange(Distance) %>% # sort by distance
  head(max_n)

head(top_n_carehomes_df)
```

#### 4.2 Obtain drivetime distances and durations through Google Maps API

Prepare data as input for querying the Google Maps API.
```{r}
# create long lat dataframe of closest n carehomes
filtered_carehome_locations <- top_n_carehomes_df %>%
    select(locationlong, locationlat) %>% 
    as.matrix()
  
  colnames(filtered_carehome_locations) <- c('lon', 'lat') # setting this consistent with mapsapi package documentation
```

Call Google distance matrix API
```{r}
# Get distance matrix from the Google Maps Distance Matrix API  
# will have <status>REQUEST_DENIED</status> if API key is not registered
doc <- mapsapi::mp_matrix(
  origins = filtered_carehome_locations,
  destinations = my_address_sf,
  mode = "driving",
  key = key,
  quiet = TRUE)

# Check status of API request.
if((xml2::xml_find_all(doc, xpath = "status") %>% xml2::xml_text()) !="OK"){
  message(xml2::xml_find_all(doc, xpath = "error_message") %>% 
                       xml2::xml_text())
}
   

```

Extract Distance matrix & Duration matrix from API return and combine them:

Distance matrix
```{r}
# Extract distances from a Google Maps Distance Matrix API response
# (wrapped into trycatch to highlight potential API key issue)
m_dist <- mapsapi::mp_get_matrix(doc, value = "distance_m")
  
colnames(m_dist) <- "driving_distance"
```

Duration matrix
```{r}
# Extract drive time duration from a Google Maps Distance Matrix API response
m_dur <- mapsapi::mp_get_matrix(doc, value = "duration_text")
colnames(m_dur) <- "driving_duration"
```

Combine outputs into df and convert meters into miles
```{r}
# create final table of closest care homes
tbl_closest_carehome <- m_dist %>% 
  tibble::as_tibble() %>% 
  cbind(driving_duration = m_dur) %>% 
  mutate(driving_distance_mi = round(driving_distance/1000/1.60934,1)) %>%  # derive distance in mi
  select(-driving_distance) %>% 
  arrange(driving_distance_mi)

knitr::kable(head(tbl_closest_carehome))

```



### 5. Mapping our results

Now that all data is in one place, we can go on and do further analysis or do some mapping...
```{r}

# set icon colour for my_location_name
icons <- awesomeIcons(
  icon = 'ios-close',
  iconColor = 'black',
  library = 'ion',
  markerColor = 'red'
)

# create output leaflet map
map <- leaflet(top_n_carehomes_df) %>% 
  addTiles() %>% 
  setView(lng = my_address_df$my_long, lat = my_address_df$my_lat, zoom = 12) %>% 
  addAwesomeMarkers(lng = my_address_df$my_long, 
                    lat = my_address_df$my_lat, 
                    icon = icons, 
                    popup = my_location_name,
                    label = my_postcode
  ) %>% 
  addMarkers(lng = ~locationlong,
             lat = ~locationlat ,
             popup = ~as.character(locationname),
             label = ~as.character(CQClocationid)
  )

map
```

