---
title: "R Code Club: Testing Code"
output: html_notebook
---

Two types of tests:  
1. Unit Tests - used in development to check that you haven't made any mistakes  
2. Run time tests - used at run time to check that the user hasn't made a mistake while running your code

Goal is to halt proceedings if something wrong occurs rather than go through potentially expensive processing.  So we need to be able to catch issues


```{r}
knitr::opts_chunk$set(chace=T, warning = F, message = F, echo = T, fig.width = 8, fig.height = 5)
```


## Basic function 1
```{r}
taking_log <- function(x) {
    return (log(x))
}
```

### Running some tests 1
```{r}
taking_log(100)
```
```{r}
taking_log("x")
```

Back to the drawing board.... Need to include some checks

## Basic function 2
```{r}
taking_log_part2 <- function(x) {
    if (is.numeric(x)) return (log(x))
    return("Cant take log")
}
```

### Running some tests 2
```{r}
taking_log_part2(100)
```
```{r}
taking_log_part2("x")
```

These work fine. Great!

## More tests ....
```{r}
taking_log_part2(-15)
```

Need to build in more tests

```{r}
taking_log_part3 <- function(x) {
    if (is.numeric(x) & x > 0 ) return (log(x))
    return("Cant take log")
}
```

## Tests now look like ..
```{r}
taking_log_part3(100)
taking_log_part3("x")
taking_log_part3(-15)
```

What about complex numbers ?

```{r}
taking_log_part3(6+5i)
```

or vectors 

```{r}
x <- taking_log_part3(c(1,-2))
x
```
Only the first element is checked -> we'd need to loop through each element or use purrr


And on it goes


## Assertive

```{r}
library(assertive)
library(magrittr)
```


```{r}
assert_is_numeric(15)
```

If passes assert, returns the value
```{r}
x <- 15 %>% 
    assert_is_numeric()
x
```

```{r}
x <- 15 %>% 
    assert_is_numeric() %>% 
    assert_all_are_greater_than(0)
x
```

Works easily with vectors
```{r}
x <- c(15,3) %>% 
    assert_is_numeric() %>% 
    assert_all_are_greater_than(0)
x
```

Otherwise crashes
```{r}
x <- c(15,-3) %>% 
    assert_is_numeric() %>% 
    assert_all_are_greater_than(0)
x
```
```{r}
x <- c(6+5i) %>% 
    assert_is_numeric() %>% 
    assert_all_are_greater_than(0)
x
```



## Assertive with Try/Catch  
We dont want to crash our code intentionally.  What do we do?


```{r}
taking_log_part5 <- function(x) {
    tryCatch(
        x %>% 
            assert_is_numeric() %>% 
            assert_all_are_greater_than(0) %>% 
            assert_is_double() %>% 
            log(),
        error=function(e){
            print(e)
            NA
        }
    )
}
```

```{r}
x <- taking_log_part5(100)
x
```

```{r}
x <- taking_log_part5("x")
x
```


```{r}
x <- taking_log_part5(-15)
x
```



```{r}
x <- taking_log_part5(6+5i)
x
```



```{r}
x <- taking_log_part5(c(100, 1000))
x
```

```{r}
x <- taking_log_part5(c(100, -15))
x
```


```{r}
x <- taking_log_part5(c(100, "x"))
x
```


# With dataframes
We can make use of the assert package functions, to check our data ....

```{r}
x <- data.frame("A"=c(1,2,3,4,5,5), "B"=c("a", "b", "c", "d", "e", "e"))

a <- x %>% 
    has_duplicates()
a
```

[Assertive documentation](https://www.google.com/url?sa=t&rct=j&q=&esrc=s&source=web&cd=&cad=rja&uact=8&ved=2ahUKEwiakaSVsJDrAhUOQxUIHQOzDwIQFjACegQIAxAB&url=https%3A%2F%2Fwww.rdocumentation.org%2Fpackages%2Fassertive%2Fversions%2F0.3-6&usg=AOvVaw0zjrQLzY6TucaB4_KXMqNB)


## Exercise
Write a function using assertive package to calculate Geometric Mean
$$\hat{x} = \sqrt[\leftroot{-3}\uproot{3} n ]{\prod_{i=1}^{n} x_i } $$

```{r}
geometric_mean <- function(x, na.rm=FALSE){
    return( exp(mean(log(x), na.rm=na.rm)) )
}
```




## A solution

```{r}
geometric_mean <- function(x, na.rm=FALSE){
    tryCatch(
        x %>% 
            assert_is_numeric() %>% 
            assert_all_are_greater_than(0) %>% 
            assert_is_double() %>% 
            log() %>% 
            mean(na.rm=na.rm) %>% 
            exp(),
        error=function(e){
            print("x makes no sense")
            NA
        }
    )
}
```

geometric_mean(c(1,1))
geometric_mean(c(2,2))
geometric_mean(c(2,x))

