---
author: "Katie Davidson [&lt;katie.davidson@dhsc.gov.uk&gt;](mailto:katie.davidson@dhsc.gov.uk)"
date: "\\25th January 2019"
output: html_document
---
  
```{r "setup", include=FALSE}
knitr::opts_knit$set(root.dir = "~/Health_R_coding_club/week2")
```

#Health R coding club
###Who has the best taste in music?
####Exploring different countries top50 playlists and their distinguishing/common features.

##### Check out the [AccessingAPI file](http://htmlpreview.github.io/?https://github.com/DataS-DH/R-coding-club/blob/master/Week2/AccessingAPI.html) if you want to try using the Spotify API yourself.

---

Set up our environment with the right packages.
```{r, eval=FALSE}
install.packages('spotifyr')
install.packages('kableExtra')
```

```{r, results='hide', message=FALSE}
library(spotifyr)
library(tidyverse)
library(kableExtra)
```
  
You will first need to load in the data. *Remember to first set your working directory to where you have saved the data.*
  
```{r, message=FALSE}
#setwd("folderLocationOfData")
top50 <- read_csv("top50playlists.csv")
```
#Lets take a quick look at the table, using the same kable foramt we used last time.
```{r}
knitr::kable(top50) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), font_size = 12) %>%
scroll_box(width = "720px", height = "500px")
```

#Lets do some quick transforming and cleaning.
#First lets convert the relevant variables to factors.
```{r}
top50$playlist_name <- as.factor(top50$playlist_name)
top50$artist_name <- as.factor(top50$artist_name)
top50$track_name <- as.factor(top50$track_name)
top50 <- top50[,-c(2,7)]
```

*You may notice we are using the knitr package above and dplyr package below but didn't install these when loading in our packages, this is because we loaded in the tidyverse package which installs a whole suite of packages including these ones, this is a great way to make sure you have everything you need if you are going to be using tidy coding approaches.*

Now the playlist_name variable is a factor we can easily see the different playlists in the dataset using levels

```{r}
levels(top50$playlist_name)
```

Lets take a look at which tracks and artists are most popular across the different playlists

```{r}
pop_tracks <- top50 %>% dplyr::count(track_name, sort=TRUE)
pop_tracks$artist_name <- top50$artist_name[match(pop_tracks$track_name, top50$track_name)]
pop_tracks

top50 %>% dplyr::count(artist_name, sort=TRUE)
```

Before we start the clustering lets clean the playlist_name column so we have a nicer shorter name for identifying a country

```{r}
top50$country <- gsub("\\-.*","", top50$playlist_name)
levels(as.factor(top50$country))
top50$country <- gsub("TOP HITS ","",top50$country)
levels(as.factor(top50$country))
top50$country <- gsub(" ","",top50$country)
levels(as.factor(top50$country))
top50$country <- as.factor(top50$country)
```
To keep things tidy lets also make the global and paraguay names match and drop the extra Japan playlist

```{r}
top50$country <- gsub("GLOBALTOP50","GLOBAL",top50$country)
top50$country <- gsub("Top50SongsinParaguay","PARAGUAY",top50$country)
top50 <- subset(top50, country!="TOPSONGSINJAPAN2019")
top50$country <- as.factor(top50$country)
```

Lets firstly compare the country based on their averages for the different features.
```{r}
top50_agg <- aggregate(top50[, c(6:17)], list(top50$country), mean)
colnames(top50_agg)[1] <- "country"
```

```{r}
knitr::kable(top50_agg) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), font_size = 12) %>%
  scroll_box(width = "720px", height = "500px")
```

We can see that key and mode are character variables so the numeric operation mean doesn't work. 
*This is why its always worth looking at your data before doing a transformation*  

Another good way to spot this is by using the summary function we looked at last week.
```{r}
summary(top50_agg)
```

We could transform these to be numberic but for now we drop them.
```{r}
top50$key <- as.factor(top50$key)
top50 %>% dplyr::count(key, sort=TRUE)
top50$mode <- as.factor(top50$mode)
top50 %>% dplyr::count(mode, sort=TRUE)
```

```{r}
top50_agg <- aggregate(top50[, c(6,7,9,11:17)], list(top50$country), mean)
colnames(top50_agg)[1] <- "country"
str(top50_agg)
```

A quick check shows all the remaining variables are numeric and have values.

```{r}
knitr::kable(top50_agg) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), font_size = 12) %>%
  scroll_box(width = "720px", height = "500px")
```

Now like last week lets normalise the data
```{r}
top50_agg_norm <- cbind(top50_agg[1],apply(top50_agg[-1],2,
                                          function(x){(x-min(x))/diff(range(x))}))
```

Again lets plot a radar chart to have a quick look at the data (this is just small tweaks to last weeks code)
```{r}
#install.packages("radarchart")
library(radarchart)
top50_agg_10 <- top50_agg[top50_agg$country %in% c("GLOBAL", "ARGENTINA", "BRAZIL", "CANADA", "FRANCE", "GERMANY", "HONGKONG", "SPAIN", "TAIWAN", "UNITEDKINGDOM", "UNITEDSTATES"),]

top50_agg_norm_10 <- cbind(top50_agg_10[1], 
                                  apply(top50_agg_10[-1],2,
                                        function(x){(x-min(x)) / diff(range(x))})) 

radarDF <- gather(top50_agg_norm_10, key=Attribute, value=Score, -country) %>%
  spread(key=country, value=Score)

chartJSRadar(scores = radarDF,
             scaleStartValue = -1, 
             maxScale =1, 
             showToolTipLabel = TRUE)
```

#####Now on to the clustering!

Before we get started we first want to rescale the numeric variables
```{r}
scaledFeatures <- scale(top50_agg[-1])
rownames(scaledFeatures) <- top50_agg$country


knitr::kable(scaledFeatures) %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), font_size = 12) %>%
  scroll_box(width = "720px", height = "500px")
```

```{r}
#install.packages("ggfortify")
library(ggfortify)
set.seed(345)
k_means <- kmeans(scaledFeatures, 3)
kmeans_plot <- autoplot(k_means,
                        main= "K-means clustering",
                        data = scaledFeatures,
                        loadings = TRUE, loadings.colour = "#CC0000",loadings.label.colour = "#CC0000",
                        loadings.label = TRUE, loadings.label.size = 3,  loadings.label.repel=T,
                        label.size = 3, label.repel = T) + scale_fill_manual(values = c("#000066","#9999CC", "#66CC99"))+ scale_color_manual(values = c("#000066", "#9999CC", "#66CC99")) + theme(plot.title=element_text(size=18, face="bold"))
```


This looks a bit messy, as its so topical lets restrict it to EU countries and rerun.
```{r}
scaledFeaturesEU <- scaledFeatures[c("AUSTRIA", "BELGIUM", "BULGARIA", "CYPRUS", "CZECHREPUBLIC", "DENMARK", "ESTONIA", "FINLAND", "FRANCE", "GERMANY", "GREECE", "HUNGARY", "IRELAND", "ITALY", "LATVIA", "LITHUANIA", "LUXEMBOURG", "MALTA", "NETHERLANDS", "POLAND", "PORTUGAL", "ROMANIA", "SLOVAKIA", "SPAIN", "SWEDEN", "UNITEDKINGDOM"),]

set.seed(345)
k_meansEU <- kmeans(scaledFeaturesEU, 3)
kmeans_plotEU <- autoplot(k_meansEU,
                        main= "K-means clustering",
                        data = scaledFeaturesEU,
                        loadings = TRUE, loadings.colour = "#CC0000",loadings.label.colour = "#CC0000",
                        loadings.label = TRUE, loadings.label.size = 3,  loadings.label.repel=T,
                        label.size = 3, label.repel = T) + scale_fill_manual(values = c("#000066","#9999CC", "#66CC99"))+ scale_color_manual(values = c("#000066", "#9999CC", "#66CC99")) + theme(plot.title=element_text(size=18, face="bold"))
```
We can tweak this to form 4 clusters, by changing the centers argument for kmeans. 
*note you will need to remember to add an extra colour to scale_fill_manual and scale_colour_manual to see these.*
```{r}
set.seed(345)
k_meansEU4 <- kmeans(scaledFeaturesEU, 4)
kmeans_plotEU4 <- autoplot(k_meansEU4,
                        main= "K-means clustering",
                        data = scaledFeaturesEU,
                        loadings = TRUE, loadings.colour = "#CC0000",loadings.label.colour = "#CC0000",
                        loadings.label = TRUE, loadings.label.size = 3,  loadings.label.repel=T,
                        label.size = 3, label.repel = T) + scale_fill_manual(values = c("#000066","#9999CC", "#66CC99", "#e7e247"))+ scale_color_manual(values = c("#000066", "#9999CC", "#66CC99", "#e7e247")) + theme(plot.title=element_text(size=18, face="bold"))
```

```{r}
k_means4 <- kmeans(scaledFeatures, 4)
kmeans_plot4 <- autoplot(k_means4,
                        main= "K-means clustering",
                        data = scaledFeatures,
                        loadings = TRUE, loadings.colour = "#CC0000",loadings.label.colour = "#CC0000",
                        loadings.label = TRUE, loadings.label.size = 3,  loadings.label.repel=T,
                        label.size = 3, label.repel = T) + scale_fill_manual(values = c("#000066","#9999CC", "#66CC99"))+ scale_color_manual(values = c("#000066", "#9999CC", "#66CC99")) + theme(plot.title=element_text(size=18, face="bold"))
```

```{r}
#install.packages("googleVis")
library(googleVis)
countries_clusters <- as_tibble(rownames_to_column(as_data_frame(k_means4$cluster)))
colnames(countries_clusters) <- c("Country", "Cluster")
WorldMap=gvisGeoChart(data = countries_clusters,locationvar="Country", colorvar="Cluster",
                      options=list(projection="kavrayskiy-vii",
                                   colorAxis="{colors:['#000066', '#9999CC', '#66CC99']}"
                      ))
plot(WorldMap)
```

```{r}
countries_clusters <- as_tibble(rownames_to_column(as_data_frame(k_meansEU$cluster)))
colnames(countries_clusters) <- c("Country", "Cluster")
WorldMap=gvisGeoChart(data = countries_clusters,locationvar="Country", colorvar="Cluster",
                      options=list(projection="kavrayskiy-vii",
                                   colorAxis="{colors:['#000066', '#9999CC', '#66CC99']}",
                                   width = 400, height = 200))
plot(WorldMap)
```

Converting cluster to vector
```{r}
top50_agg_norm_83 <- cbind(top50_agg[1], 
                           apply(top50_agg[-1],2,
                                 scale)) 

library(radarchart)
library(tidyr)
cluster_centers <- as.data.frame(k_means$centers)
cluster <- c("Cluster 1", "Cluster 2", "Cluster 3")
cluster_centers <- cbind(cluster, cluster_centers)

radarDF2 <- gather(cluster_centers, key=Attribute, value=Score, -cluster) %>%
  spread(key=cluster, value=Score)
#we change the colours according to clusters
colMatrix = matrix(c(c(4,24,102), c(135,133,193), c(87,196,135)), nrow = 3)
#chart
chartJSRadar(scores = radarDF2,
             scaleStartValue = -4, 
             maxScale =1.5, 
             showToolTipLabel = TRUE,
             colMatrix = colMatrix)
```

We can further explore individual features, lets look at loudness.
```{r}
library(stringr)
clusters <- as.vector(k_means$cluster)
clusters <- str_replace_all(clusters, "1", "Cluster 1")
clusters <- str_replace_all(clusters, "2", "Cluster 2")
clusters <- str_replace_all(clusters, "3", "Cluster 3")
clusters[28] <- "Global"
```

```{r}
top50_agg_norm_83 <- cbind(top50_agg_norm_83, cluster = clusters)
loudness_subset <- top50_agg_norm_83[,c("country","loudness", "cluster")]

loudness_subset <- loudness_subset[order(loudness_subset$loudness, decreasing = TRUE), ]

loudness_plot <- ggplot(loudness_subset, aes(x = reorder(country, loudness), y = loudness, label=loudness)) + xlab("Country") + ylab("Loudness") + geom_bar(stat='identity', aes(fill=cluster), width = .5) + scale_fill_manual(name="Cluster", labels = c("Cluster 1", "Cluster 2", "Cluster 3", "Global"), values = c("Cluster 1"="#000066", "Cluster 2"="#9999CC", "Cluster 3" = "#66CC99", "Global" = "indianred2")) + labs(title="Loudness feature", subtitle= "Diverging Bars") + coord_flip()
loudness_plot
```