---
author: "Katie Davidson [&lt;katie.davidson@dhsc.gov.uk&gt;](mailto:katie.davidson@dhsc.gov.uk)"
date: "\\25th January 2019"
output: html_document
---
  
```{r "setup", include=FALSE}
knitr::opts_knit$set(root.dir = "~/Health_R_coding_club/week2")
```

```{r, eval=FALSE}
install.packages('spotifyr')
install.packages('kableExtra')
```

```{r, echo=FALSE}
library(spotifyr)
library(tidyverse)
library(kableExtra)
```
#Health R coding club
##Who has the best taste in music?
###Exploring different countries top50 playlists and there distinguishing/common features.

#### Check out the LoadingAPIdata file if you want to try using the Spotify API yourself.

---
  
  ####You will first need to load in the data. *Remember to first set your working directory to where you have saved the data.*
  
  ```{r}
#setwd("folderLocationOfData")
top50 <- read_csv("top50playlists.csv")
```

#We can do some basic 

```{r}
top50 %>% dplyr::count(track_name, sort=TRUE)
```
*You may notice we are using the dplyr package above and knitr package below but didn't install these when loading in our packages, this is because we loaded in the tidyverse package which installs a whole suite of packages including these ones, this is a great way to make sure you have everything you need if you are going to be using tidy coding approaches.*

```{r}
knitr::kable(top50) %>%
kableExtra::kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"), font_size = 12) %>%
scroll_box(width = "720px", height = "500px")
```
