---
title: 'Interactive graphs with Plotly'
author: "James Riley"
institute: "DHSC"
date: "2020-10-22"
output: slidy_presentation
---
  
  
```{r setup, include = FALSE, warning=FALSE, message=FALSE}

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)
library("tidyverse")
library("knitr")
library("magrittr")
library("downloader")
library("plotly")

```


## Overview


<iframe src="https://plotly.com/r/" width="100%" height="100%"></iframe>



## Required Packages

See requirements.R in the same folder as the slides. 

I've taken [Baby names in England & Wales 1996-2019](https://www.ons.gov.uk/peoplepopulationandcommunity/birthsdeathsandmarriages/livebirths/datasets/babynamesinenglandandwalesfrom1996) as a working dataset. I've tidied that data since we looked at data importing earlier.

I've been entirely egotistical and used "James" as an example - grab the code for this presentation and change it to your/your kids' names.

I'll generate one or two other datasets on-the-fly.

```{r}
names <- read_csv(here::here("2020-10-22 Interactive Graphs/data/babynames.csv")) %>%
  select(year, gender, Name, count)
```

## Exploring data

Traditionally:

```{r}
sine <- tibble(t = seq(0, 3, length.out = 10000),
               y = sin(880 * t))

sine %>%
  ggplot(aes(x = t, y = y)) + geom_line()
```

Erm... change the scale? Try new numbers wait...



## With Plotly

```{r}
sine %>%
  plot_ly(x = ~t, y = ~y) %>%
  add_lines()
```


## Shortcut

If you're already familiar with ggplot, there's a conversion function that usually works:

```{r}
g <- names %>%
  filter(Name == "JAMES", gender == "Boys") %>%
  ggplot(aes(x = year, y = count)) + geom_line() + theme_minimal()

ggplotly(g)
```


But it's not usually pretty...



---



Plotly is usually happier when we don't go via ggplot. Which does mean learning a new syntax:

```{r}
names %>%
  filter(Name == "JAMES", gender == "Boys") %>%
  plot_ly(x = ~year, y = ~count) %>%
    add_lines()
```




---

That makes it look like Jameses are declining to 0/year, so we can reset the y-axis:

```{r}
names %>%
  filter(Name == "JAMES", gender == "Boys") %>%
  plot_ly(x = ~year, y = ~count) %>%
    add_lines() %>%
    layout(
      yaxis = list(rangemode = "tozero")
    )

```


---

We can map variables to plot aesthetics, like in ggplot:

```{r}
names %>%
  filter(Name == "ALEX") %>%
  plot_ly(x = ~year, y = ~count, color = ~gender) %>%
  add_lines()

```

---

It's not been _very_ interactive yet. Let's animate, looking at top 10 girls/boys names by year.

```{r}
names %>%
  filter(gender == "Girls") %>%
  group_by(year) %>%
  slice_max(count, n = 10) %>%
  arrange(desc(count)) %>%
  plot_ly(x = ~Name, y = ~count, frame = ~year) %>%
    add_bars() %>%
    animation_opts(redraw = TRUE, frame = 1000, transition = 0)
```


---

```{r}
names %>%
  filter(gender == "Boys") %>%
  group_by(year) %>%
  slice_max(count, n = 10) %>%
  arrange(desc(count)) %>%
  plot_ly(x = ~Name, y = ~count, frame = ~year) %>%
    add_bars() %>%
    animation_opts(redraw = TRUE, frame = 1000, transition = 0)

```


---

You *can* make 3D graphs. You rarely *should*:

```{r}
top_10 <- names %>%
  group_by(gender, Name) %>%
  summarise(total = sum(count)) %>%
  ungroup() %>%
  slice_max(total, n = 10)

```

```{r}
names %>%
  filter(Name %in% top_10$Name) %>%
  plot_ly(x = ~year, y = ~Name, z = ~count) %>%
  add_markers()
```