---
title: "Intro to SQL workshop"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
library(dplyr)
library(DBI)
```

# Intro to SQL


First lets load some data to play with. This is a table of population data from the ONS, which we have stored in a table called ons_out.

```{r, echo=FALSE}
# Load data
ons_pop <- readxl::read_excel("ukmidyearestimates20192020ladcodes.xls", sheet = "MYE2 - Persons", skip = 3)
# Get column names for age-aggregates
over_18 <- colnames(ons_pop)[as.integer(colnames(ons_pop)) > 17 & !is.na(as.integer(colnames(ons_pop)))]
over_85 <- colnames(ons_pop)[as.integer(colnames(ons_pop)) > 84 & !is.na(as.integer(colnames(ons_pop)))]
over_65 <- colnames(ons_pop)[as.integer(colnames(ons_pop)) > 64 & !is.na(as.integer(colnames(ons_pop)))]
# Create pre-processed data
ons_out <- ons_pop %>%
  dplyr::transmute(lad_code = Code, 
                   lad_name = Name, 
                   geo_type = Geography1, 
                   all_ages = `All ages`,
                   over_18 = rowSums(.[over_18]),
                   over_85 = rowSums(.[over_85]),
                   over_65 = rowSums(.[over_65]))

print(ons_out)

```

## Creating a database

SQL stands for "Structured Query Language", and it is designed for querying tables of data stored in databases.

Typically, the database is stored in a remote server somewhere. You connect to the server, via a programming language like R or python, or through specialised software like SQL Server Management Studio. You then send a query to the server in the form of a SQL statement, and the server sends back the data you asked for.

For this workshop, the database will be stored in a file on our local machine instead. The R code below creates this database, called my-sqlite-db.sqlite, and adds the above table to it:

```{r}
mydb <- DBI::dbConnect(RSQLite::SQLite(), "my-sqlite-db.sqlite")

DBI::dbWriteTable(mydb, "ons_out", ons_out, overwrite=T)
```

**Note**, the my-sqlite-db.sqlite appears in my folder.  

If I wanted to add data to an existing table I would replace the `overwrite=T` with `append=T`

I can also list the tables in my database - I find this useful as I forget the names
```{r}
DBI::dbListTables(mydb)
```

Also, when you finish with a database you need to disconnect
```{r}
# DBI::dbDisconnect(mydb)
```

We've used the DBI package in R here to create the database. You can also use functions from this package to connect to databases, and execute SQL queries, as we'll see below. But the focus of this workshop is on SQL itself, rather than using it inside R. Whether you're using R, python, or writing queries directly in database software, the syntax of the SQL queries will usually remain the same.

We'll now review how to select the data we want from this table, using 'SELECT' queries.


## Selecting columns

In its simplest form, a SELECT query allows you to select some subset of columns from a table. It has the following format:

SELECT \<column name 1\>, \<column name 2\>, ... FROM \<table name\>

For example:

```{r}
dplyr::tibble(DBI::dbGetQuery(mydb, "
      SELECT lad_name, all_ages FROM ons_out
      "))
```

## Selecting new columns

We can also create new columns out of the columns which already exist, and select those too, all within a single SELECT query. We can apply basic arithmetic operations to columns containing numerical values. The below query creates a new column which tells us the proportion of the population who are under 18.

```{r}
dplyr::tibble(DBI::dbGetQuery(mydb, "
SELECT lad_name,
all_ages,
(all_ages - over_18) / all_ages AS prop_under_18
FROM ons_out
                    "))
```

Another really useful tool for creating new columns is "CASE WHEN". For example, the statement below creates a new column classifying each area as younger than average or older than average, based on the proportion under 18:

```{r}
dplyr::tibble(DBI::dbGetQuery(mydb, "
SELECT lad_name,
all_ages,
(all_ages - over_18) / all_ages AS prop_under_18,
CASE WHEN (all_ages - over_18) / all_ages < 0.219 THEN 'Older than average'
WHEN (all_ages - over_18) / all_ages > 0.221 THEN 'Younger than average'
ELSE 'Average' END AS age_classification
FROM ons_out
                    "))
```

Finally, we can select all existing columns in the table using *:

```{r}
dplyr::tibble(DBI::dbGetQuery(mydb, "
SELECT *,
(all_ages - over_18) / all_ages AS prop_under_18,
CASE WHEN (all_ages - over_18) / all_ages < 0.219 THEN 'Older than average'
WHEN (all_ages - over_18) / all_ages > 0.221 THEN 'Younger than average'
ELSE 'Average' END AS age_classification
FROM ons_out
                    "))
```


## Selecting rows

Suppose we don't want all the rows in the table, but only those which satisfy certain criteria. We can do this using a WHERE clause, which goes after the table name. The below SQL query only returns rows which correspond to regions.

```{r}
dplyr::tibble(DBI::dbGetQuery(mydb, "
SELECT lad_name,
all_ages,
(all_ages - over_18) / all_ages AS prop_under_18
FROM ons_out
WHERE geo_type = 'Region'
                    "))
```


Some basic operations you can use in a WHERE clause are given below:

= equals

\<\> not equal

\< less than

\> greater than

NOT not

You can also combine conditions using AND and OR. For example, suppose we want all the rows below the level of country and region:

```{r}
dplyr::tibble(DBI::dbGetQuery(mydb, "
SELECT lad_name,
all_ages,
(all_ages - over_18) / all_ages AS prop_under_18
FROM ons_out
WHERE NOT (geo_type = 'Region' OR geo_type = 'Country')
                    "))
```

A slightly more complicated relation that you can use for comparing strings is LIKE. The '%' symbol can be used as a wildcard. For example, if for some reason we wanted to know the population statistics for all local authorities which begin with 'H':

```{r}
dplyr::tibble(DBI::dbGetQuery(mydb, "
SELECT lad_name,
all_ages,
(all_ages - over_18) / all_ages AS prop_under_18
FROM ons_out
WHERE lad_name LIKE 'H%'
                    "))
```

**Caution:** whether or not string comparisons are case sensitive is not standard. It can vary by the type of database. For the SQLite database being used here, the below query returns no results:

```{r}
dplyr::tibble(DBI::dbGetQuery(mydb, "
SELECT lad_name,
all_ages,
(all_ages - over_18) / all_ages AS prop_under_18
FROM ons_out
WHERE geo_type = 'rEgIon'
                    "))
```

But if this database were stored on Microsoft SQL Server, the above query would still select the rows of geo_type 'Region'.

## Sorting the output

Suppose we want to list the regions in decreasing order of their proportion of 18 year olds, and then use population to break ties (higher populations placed higher). We can do this using ORDER BY:

```{r}
dplyr::tibble(DBI::dbGetQuery(mydb, "
SELECT lad_name,
all_ages,
(all_ages - over_18) / all_ages AS prop_under_18
FROM ons_out
WHERE geo_type = 'Region'
ORDER BY prop_under_18 desc, all_ages
                    "))
```

## Joining tables together

Below is a second table, le_all, which contains life expectancy statistics at an LA level.

```{r, echo=FALSE}
load_le <- function(sheet_name){
  # Read data, remove empty rows
  le <- readxl::read_excel("leatbirthandatage65byukla201618.xlsx", sheet = sheet_name, skip= 2) %>%
    dplyr::filter(!is.na(`Area Codes`))
  # Set column names
  colnames(le) <- c(c('lad_code', 'area_name', 'county', 'district'), colnames(le)[5:ncol(le)])
  # Fill Local Authority names into single column
  tidied_le <- le %>%
    tidyr::pivot_longer(cols = -(lad_code:district),
                        names_to = 'year',
                        values_to = 'le') %>%
    dplyr::mutate(lad_name = ifelse(is.na(area_name) & is.na(county), district,
                                    ifelse(is.na(area_name), county,
                                           area_name))) %>%
    # Only keep latest year
    dplyr::filter(year == max(year)) %>%
    # Pivot back to a single column, named based on sheet name
    tidyr::pivot_wider(id_cols = c(lad_code, lad_name),
                       names_from = year,
                       names_prefix = paste0(gsub(' ', '_', sheet_name), '_'),
                       values_from = le)
  return(tidied_le)
}
le_males <- load_le("LE at age 65 - Males")
le_females <- load_le("LE at age 65 - Females")

le_all <- dplyr::full_join(le_males, le_females, by = c("lad_code", "lad_name"))

colnames(le_all) <- c("lad_code", "lad_name", "le_at_65_males", "le_at_65_females")

print(le_all)
```

Lets start by adding this table to our database:

```{r}
DBI::dbWriteTable(mydb, "le_all", le_all, overwrite=T)
```


Suppose we want to join this table to our original table, to create a single table with both population and life expectancy data for each LA. This is straightforward to do in SQL. We just need to specify a condition for when to match rows together. Here, we can match on the lad_code columns in each table.

```{r}
dplyr::tibble(DBI::dbGetQuery(mydb, "
SELECT ons_out.lad_name, all_ages, le_at_65_males, le_at_65_females
FROM ons_out INNER JOIN le_all ON ons_out.lad_code = le_all.lad_code
                    "))
```


Note that because "lad_name" is a column name in both tables, we need to specify which table we mean when we refer to that column in the select query.

If a lad_code had appeared multiple times in either table, every possible match will be returned in the joined tables.

Also note that here we used an INNER JOIN. There are actually 4 different kinds of join that you can choose from.

### INNER JOIN

With an INNER JOIN, if a row in one table has no match in the other table, then it will be excluded from the results. In this case, Great Britain did not appear as a row in the life expectancy table, so it is missing above, as we can verify by adding a WHERE clause:

```{r}
dplyr::tibble(DBI::dbGetQuery(mydb, "
SELECT ons_out.lad_name, all_ages, le_at_65_males, le_at_65_females
FROM ons_out INNER JOIN le_all ON ons_out.lad_code = le_all.lad_code
WHERE ons_out.lad_name = 'GREAT BRITAIN'
                    "))
```

### LEFT JOIN

With a LEFT JOIN, all rows in the left table will be included in the results. If a row in the left table has no match in the right table, then missing values will be entered in the corresponding columns. However, with a LEFT JOIN, rows in the right table with no match in the left table are still excluded. Below, we do a left join with ons_out on the left, and a left join with ons_out on the right, to see that Great Britain is included when ons_out is on the left but still excluded otherwise:

```{r}
dplyr::tibble(DBI::dbGetQuery(mydb, "
SELECT ons_out.lad_name, all_ages, le_at_65_males, le_at_65_females
FROM ons_out LEFT JOIN le_all ON ons_out.lad_code = le_all.lad_code
WHERE ons_out.lad_name = 'GREAT BRITAIN'
                    "))
```

Note that although GREAT BRITAIN now appears, its life expectancy stats are missing, because there was no GREAT BRITAIN entry in the life expectancy table.

```{r}
dplyr::tibble(DBI::dbGetQuery(mydb, "
SELECT ons_out.lad_name, all_ages, le_at_65_males, le_at_65_females
FROM le_all LEFT JOIN ons_out ON ons_out.lad_code = le_all.lad_code
WHERE ons_out.lad_name = 'GREAT BRITAIN'
                    "))
```

If ons_out is on the right, then GREAT BRITAIN is still excluded when we do a LEFT JOIN, as it has no match in the left hand table.

### RIGHT JOIN

A RIGHT JOIN is the opposite of a LEFT JOIN. All rows in the right hand table are included in the joined table, but rows in the left hand table with no match are excluded.

### FULL OUTER JOIN

All rows in both tables are included.

## Applying operations over multiple rows

We have already seen how to perform row by row operations, when we computed a new column containing the proportion of the population who are under 18. But what if we want to perform some operation involving multiple rows? SQL has several inbuilt functions which allow us to do this.

Suppose we want to know the total number of people who live in metropolitan districts. We can use the WHERE clause to only return rows corresponding to metropolitan districts, and use the sum(.) function to add up all the rows:

```{r}
dplyr::tibble(DBI::dbGetQuery(mydb, "
SELECT sum(all_ages) AS total_pop_in_met_districts
FROM ons_out WHERE geo_type = 'Metropolitan District'
                    "))
```

Other functions you can apply here include mean(.), max(.), min(.).

But what if we want to apply the operation not to all rows, but within certain groups of rows? For example, suppose I want to take all of the sub-national and sub-regional geo_types, and for each one, compute the proportion of the population below 18. We can do that using GROUP BY, which ensures that the operations will be applied within each group:

```{r}
dplyr::tibble(DBI::dbGetQuery(mydb, "
SELECT geo_type, sum(all_ages) AS population,
(sum(all_ages) - sum(over_18)) / sum(all_ages) AS prop_below_18
FROM ons_out
WHERE NOT (geo_type = 'Region' OR geo_type = 'Country')
GROUP BY geo_type
                    "))
```

The sum operation is now applied within each group, instead of over every row.

Finally, suppose we wanted to only include geo_types with more than 2 million people living in them. Unfortunately, in SQL you cannot include aggregation functions like sum(.) inside the WHERE clause (the WHERE clause is used to select the rows of the table before the grouping is applied). If you want to do an additional selection on your final output, you can use HAVING. This works just like WHERE, and goes at the end. For example:

```{r}
dplyr::tibble(DBI::dbGetQuery(mydb, "
SELECT geo_type, sum(all_ages) AS population,
(sum(all_ages) - sum(over_18)) / sum(all_ages) AS prop_below_18
FROM ons_out
WHERE NOT (geo_type = 'Region' OR geo_type = 'Country')
GROUP BY geo_type
HAVING population > 2000000
                    "))
```


## Putting it all together

The below query shows the total population living under each geography type, the proportion of that population that is under 18, and the life expectancy of a 65 year old male or 65 year old female living in that geography type (which needs to be computed by a weighted average), where the total population in the geography type is more than 2 million, and sorted in decreasing order of their percentage of under 18s. Strictly speaking, we should use the population of 65 year olds in each area to weight the average, but here we weight by total population for simplicity.

```{r}
dplyr::tibble(DBI::dbGetQuery(mydb, "
SELECT geo_type,
sum(all_ages) AS population,
(sum(all_ages) - sum(over_18)) / sum(all_ages) AS prop_below_18,
sum(all_ages * le_at_65_males) / sum(all_ages) AS av_le_at_65_males,
sum(all_ages * le_at_65_females) / sum(all_ages) AS av_le_at_65_females
FROM ons_out LEFT JOIN le_all ON ons_out.lad_code = le_all.lad_code
WHERE NOT (geo_type = 'Region' OR geo_type = 'Country')
GROUP BY geo_type
HAVING population > 2000000
ORDER BY prop_below_18 desc
                    "))
```

Note that metropolitan districts are clearly contained within metropolitan counties, and the life expectancy for females will be slightly off because we used total population instead of population of 65 year olds to weight the average.

### UNION

UNION lets you combine the rows from two separate tables into one larger table (the equivalent of rbind in R).  I'll just split the data into two tables, add them to the database and then UNION them to get the original table

```{r}
table1 <- dbGetQuery(mydb, "SELECT * FROM ons_out WHERE rowid  <= 214") 
dbWriteTable(mydb, "Table1", table1, overwrite=T)

table2 <- dbGetQuery(mydb, "SELECT * FROM ons_out WHERE rowid  > 214") 
dbWriteTable(mydb, "Table2", table2, overwrite=T)
```

These tables now appear in the database
```{r}
dbListTables(mydb)
```


```{r}
table3 <- dbGetQuery(mydb, 
"SELECT * FROM Table1 
UNION ALL
SELECT * FROM Table2")
```

```{r}
dim(table3)
```


## Dynamic SQL

In this workshop, we have been running all of our SQL queries from inside R. This is not necessarily how you would normally use SQL. If you are working with a database, you will typically have access to some software which lets you enter SQL queries directly and execute them.

However, there is an advantage to running SQL queries from inside a programming language like R or python. Because R views the SQL queries as just ordinary strings, we can manipulate them like strings, and generate our queries dynamically.

For example, suppose we wanted to create 3 population tables: one for countries, one for regions, and one for unitary authorities. If you were writing raw SQL code, the simplest way to do this would probably be to write out 3 full SQL queries, one for each table, differing only in the name of the geo_type being selected. But if we're running our SQL queries from inside R, we can do something like this:

```{r}
for (g_type in c("Country", "Region", "Unitary Authority")) {
  print(dplyr::tibble(DBI::dbGetQuery(mydb, paste0("
                                    SELECT lad_name, all_ages
                                    FROM ons_out WHERE geo_type = '", g_type, "'
                                    ORDER BY all_ages desc
                                  "))))
}
```


```{r, echo=FALSE}
DBI::dbDisconnect(mydb)
file.remove("my-sqlite-db.sqlite")
```